---
title: "Small RNAseq: Differential Expression Analysis"
date: "`r Sys.Date()`"
author:
  - name: Arun Seetharam
    affiliation: Tuteja Lab
    affiliation_url: https://www.tutejalab.org
output:
  #pdf_document: default
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
options(max.print = "125")
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "assets/",
  fig.width = 8,
  prompt = FALSE,
  tidy = FALSE,
  message = FALSE,
  warning = TRUE
)
knitr::opts_knit$set(width = 75)
```
# Environment Setup

```{bash eval = FALSE}
salloc -N 1 --exclusive -p amd -t 8:00:00
conda env create -f conda-env.yml
conda activate smallrna
```

# Downloading datasets

## Raw data

Raw data was downloaded from the sequencing facility using the secure link, with `wget` command. The downloaded files were checked for md5sum and compared against list of files expected as per the input samples provided.

```{bash, eval=FALSE, engine="sh"}
wget https://oc1.rnet.missouri.edu/xyxz
# link masked 
# GEO link will be included later
# merge files of same samples (technical replicates)
paste <(ls *_L001_R1_001.fastq.gz) <(ls *_L002_R1_001.fastq.gz) | \
   sed 's/\t/ /g' |\
   awk '{print "cat",$1,$2" > "$1}' |\
   sed 's/_L001_R1_001.fastq.gz/.fq.gz/2' > concatenate.sh
chmod +x concatenate.sh
sh concatenate.sh
```


## Genome/annotation

Additional files required for the analyses were downloaded from [GenCode](https://www.gencodegenes.org/mouse/). The downloaded files are as follows:

```{bash eval = FALSE} 
wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M30/GRCm39.primary_assembly.genome.fa.gz
wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M30/gencode.vM30.annotation.gff3.gz
wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M30/gencode.vM30.annotation.gtf.gz
gunzip GRCm39.primary_assembly.genome.fa.gz
gunzip gencode.vM30.annotation.gff3.gz
gunzip gencode.vM30.annotation.gtf.gz
```


## FastQC (before processing)

```{bash eval = FALSE} 
for fq in *.fq.gz; do
  fastqc --threads $SLURM_JOB_CPUS_PER_NODE $fq;
done
mkdir -p fastqc_pre
mv *.zip *.html fastqc_pre/
```


# Mapping

To index the genome, following command was run (in an interactive session).

```{bash eval = FALSE} 
fastaGenome="GRCm39.genome.fa"
gtf="gencode.vM30.annotation.gtf"
STAR --runThreadN $SLURM_JOB_CPUS_PER_NODE \
     --runMode genomeGenerate \
     --genomeDir $(pwd) \
     --genomeFastaFiles $fastaGenome \
     --sjdbGTFfile $gtf \
     --sjdbOverhang 1
```
Each `fastq` file was mapped to the indexed genome as using `runSTAR_map.sh` script shown below:

```{bash eval = FALSE}
#!/bin/bash
read1=$1
out=$(basename ${read1%%.*})
STARgenomeDir=$(pwd)
# illumina adapter
adapterseq="AGATCGGAAGAGC"
STAR \
    --genomeDir ${STARgenomeDir} \
    --readFilesIn ${read1} \
    --outSAMunmapped Within \
    --readFilesCommand zcat \
    --outSAMtype BAM SortedByCoordinate \
    --quantMode GeneCounts \
    --outFilterMultimapNmax 20 \
    --clip3pAdapterSeq ${adapterseq} \
    --clip3pAdapterMMp 0.1 \
    --outFilterMismatchNoverLmax 0.03 \
    --outFilterScoreMinOverLread 0 \
    --outFilterMatchNminOverLread 0 \
    --outFilterMatchNmin 16 \
    --outFileNamePrefix ${out} \
    --alignSJDBoverhangMin 1000 \ 
    --alignIntronMax 1 \
    --runThreadN ${SLURM_JOB_CPUS_PER_NODE} \
    --genomeLoad LoadAndKeep \
    --limitBAMsortRAM 30000000000 \
    --outSAMheaderHD "@HD VN:1.4 SO:coordinate"
```

Mapping was run with a simple loop:

```{bash eval = FALSE}
for fq in *.fq.gz; do
  runSTAR_map.sh $fq;
done
```


# Counting Stats

Counts generated by `STAR` with option `--quantMode GeneCounts` were parsed to generate summary stats as well as to extract 
annotated small RNA feature counts.


```{bash eval = FALSE}
mkdir -p counts_files
# copy counts for each sample
cp *ReadsPerGene.out.tab counts_files/
cd counts_files
# merge counts
join_files.sh *ReadsPerGene.out.tab |\
   sed 's/ReadsPerGene.out.tab//g' |\
   grep -v "^N_" > counts_star.tsv
# merge stats
join_files.sh *ReadsPerGene.out.tab |\
   sed 's/ReadsPerGene.out.tab//g' |\
   head -n 1 > summary_star.tsv
join_files.sh *ReadsPerGene.out.tab |\
   sed 's/ReadsPerGene.out.tab//g' |\
   grep "^N_" >> summary_star.tsv
# parse GTF to extact gene.id and its biotype:
gtf=gencode.vM30.annotation.gtf
awk 'BEGIN{OFS=FS="\t"} $3=="gene" {split($9,a,";"); print a[1],a[2]}' ${gtf} |\
   awk '{print $4"\t"$2}' |\
   sed 's/"//g' > GeneType_GeneID.tsv
cut -f 1 GeneType_GeneID.tsv | sort |uniq > features.txt
```

The information for `biotype` as provided by the [`gencodegenes`](https://www.gencodegenes.org/pages/biotypes.html) were used 
for categorizing `biotype`.

The `smallRNA` group consists of following `biotype`:

```
miRNA
misc_RNA
scRNA
snRNA
snoRNA
sRNA
scaRNA
```
The full table is as follows:

```{r table1, warnings=TRUE, message=FALSE, results='asis'}
library(knitr)
setwd("/work/LAS/geetu-lab/arnstrm/mouse.trophoblast.smallRNAseq")
file1="assets/GeneType_Group.tsv"
info <-
  read.csv(
    file1,
    header = TRUE,
    sep = "\t",
    stringsAsFactors = TRUE
  )
kable(info, caption = "Table 1: biotype and its groupings")
```

A samples table (`samples.tsv`) categorizing samples to its condition were also generated:

```{r table3, warnings=TRUE, message=FALSE, results='asis'}
file2="assets/samples.tsv"
samples <-
  read.csv(
    file2,
    header = TRUE,
    sep = "\t",
    stringsAsFactors = TRUE
  )
kable(samples, caption = "Table 2: Samples in the study")
```

This information was then merged withe counts table to generate QC plots:

```{bash eval = FALSE}
awk 'BEGIN{OFS=FS="\t"}FNR==NR{a[$1]=$2;next}{ print $2,$1,a[$1]}' \
    GeneType_Group.tsv GeneType_GeneID.tsv > GeneID_GeneType_Group.tsv

awk 'BEGIN{OFS=FS="\t"}FNR==NR{a[$1]=$2"\t"$3;next}{print $1,a[$1],$0}' \
    GeneID_GeneType_Group.tsv counts_star.tsv |\
    cut -f 1-3,5- > processed_counts_star.tsv
```


### Plotting the mapping summary and count statistics for various biotypes:

```{r libload, warnings=TRUE, message=FALSE}
library(scales)
library(tidyverse)
library(plotly)
```

```{r dataload, warnings=TRUE, message=FALSE}
setwd("/work/LAS/geetu-lab/arnstrm/mouse.trophoblast.smallRNAseq")
file1="assets/processed_counts_star.tsv"
file2="assets/summary_stats_star.tsv"
counts <-
  read.csv(
    file1,
    sep = "\t",
    stringsAsFactors = TRUE
  )
subread <-
  read.csv(
    file2,
    sep = "\t",
    stringsAsFactors = TRUE
  )
# convert long format
counts.long <- gather(counts, Sample, Count, Dif_D6_1_S4:Undif_D2_4_S5, factor_key=TRUE)
subread.long <- gather(subread, Sample,  Count, Dif_D6_1_S4:Undif_D2_4_S5, factor_key=TRUE)
# organize
counts.long$Group <-
  factor(
    counts.long$Group,
    levels = c(
      "coding_genes",
      "non_conding_RNA",
      "long_non_conding_RNA",
      "pseudogenes",
      "unconfirmed_genes",
      "Ig_genes"
    )
  )

subread.long$Assignments <-
  factor(
    subread.long$Assignments,
    levels = c(
      "N_input",
      "N_unmapped",
      "N_multimapping",
      "N_unique",
      "N_ambiguous",
      "N_noFeature"
    )
  )
```



```{r readstat1, fig.cap="Figure 1: STAR read mapping and feature assignment. Here, `N_input` is total input reads, `N_unmapped` is reads that were either too short to map after adapter removal or had higher mismatch rate to place reliably on the genome, `N_multimapping` is reads mapped to multiple loci, `N_unique` is reads mapped to unique loci. A subset of `N_unique` reads that were unable to clearly assign to a feature or assign any feature at all are grouped as `N_ambigious` or `N_noFeature`, respectively", fig.width=9, fig.height=7}
ggplot(subread.long, aes(x = Assignments, y = Count, fill = Assignments)) +
  geom_bar(stat = 'identity') +
  labs(x = "Subread assingments", y = "reads") + theme_minimal() +
  scale_y_continuous(labels = label_comma()) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1,
      size = 12
      ),
    strip.text = element_text(
      face = "bold",
      color = "gray35",
      hjust = 0,
      size = 10
    ),
    strip.background = element_rect(fill = "white", linetype = "blank"),
    legend.position = "none"
  ) +
  facet_wrap("Sample", scales = "free_y", ncol = 4) 
```

```{r readstat2, fig.cap="Figure 2: Features with read counts", fig.width=9, fig.height=7}
g <- ggplot(counts.long, aes(x = Group, y = Count, fill = Group)) +
  geom_bar(stat = 'sum') +
  labs(x = "biotype", y = "read counts") + theme_minimal() +
  scale_y_continuous(labels = label_comma()) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1,
      size = 12
    ),
    strip.text = element_text(
      face = "bold",
      color = "gray35",
      hjust = 0,
      size = 10
    ),
    strip.background = element_rect(fill = "white", linetype = "blank"),
    legend.position = "none"
  ) +
  facet_wrap("Sample", scales = "free_y", ncol = 4)
#ggplotly(g)
g
```


```{r readstat3, fig.cap="Figure 3: non-coding biotype read counts", fig.width=12, fig.height=6}
counts.nc <- filter(counts.long, Group %in% "non_conding_RNA" )
counts.nc$GeneType <-
  factor(
    counts.nc$GeneType,
    levels = c(
      "miRNA",
      "misc_RNA",
      "snoRNA",
      "snRNA",
      "sRNA",
      "scRNA",
      "scaRNA",
      "Mt_tRNA",
      "Mt_rRNA",
      "rRNA",
      "ribozyme"
    )
  )

g <- ggplot(counts.nc, aes(x = GeneType, y = Count, fill = GeneType)) +
  geom_bar(stat = 'sum') +
  labs(x = "biotype", y = "read counts") + theme_minimal() +
  scale_y_continuous(labels = label_comma()) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1,
      size = 12
    ),
    strip.text = element_text(
      face = "bold",
      color = "gray35",
      hjust = 0,
      size = 10
    ),
    strip.background = element_rect(fill = "white", linetype = "blank"),
    legend.position = "none"
  ) +
  facet_wrap("Sample", scales = "free_y", ncol = 4)
#ggplotly(g)
g
```


subset the counts file to select only smallRNA genes

```{r subset, warnings=TRUE, message=FALSE}
snrna <- c('miRNA',
           'misc_RNA',
           'scRNA',
           'snRNA',
           'snoRNA',
           'sRNA',
           'scaRNA')
cts <- filter(counts, GeneType %in% snrna) %>%
  select(Geneid, Dif_D6_1_S4:Undif_D2_4_S5)
write_delim(cts, file = "assets/noncoding_counts_star.tsv", delim = "\t")
```


This `noncoding_counts_star.tsv` and `samples.tsv` file will be used for `DESeq2` analyses.


# DESeq2

For the next steps, we used `DESeq2` for performing the DE analyses. Results were visualized as volcano plots and tables were exported to excel.


## Load packages

```{r, warnings=TRUE, message=FALSE}
setwd("/work/LAS/geetu-lab/arnstrm/mouse.trophoblast.smallRNAseq")
library(DESeq2)
library(RColorBrewer)
library(pheatmap)
library(genefilter)
library(ggrepel)
library(biomaRt)
library(reshape2)
library(PupillometryR)
```

## Import counts and sample metadata


The `counts` data and its associated metadata (`coldata`) are imported for analyses.

```{r dataset, warnings=TRUE, message=FALSE}
counts = 'assets/noncoding_counts_star.tsv'
groupFile = 'assets/samples.tsv'
coldata <-
  read.csv(
    groupFile,
    row.names = 1,
    sep = "\t",
    stringsAsFactors = TRUE
  )
cts <- as.matrix(read.csv(counts, sep = "\t", row.names = "Geneid"))

```

  <!-- Inspect the `cts` data. -->
  
  <!-- ```{r ctscheck} -->
  <!-- DT::datatable(cts) -->
  <!-- ``` -->
Reorder columns of `cts` according to `coldata` rows. Check if samples in both files match.

```{r order, warnings=TRUE, message=FALSE}
colnames(cts)
all(rownames(coldata) %in% colnames(cts))
cts <- cts[, rownames(coldata)]
```

## Normalize

The batch corrected read counts are then used for running DESeq2 analyses

```{r deseq2, warnings=TRUE, message=FALSE}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ Group)
vsd <- vst(dds, blind = FALSE, nsub =500)
keep <- rowSums(counts(dds)) >= 5
dds <- dds[keep, ]
dds <- DESeq(dds)
dds
```


```{r vst, warnings=TRUE, message=FALSE}
vst <- assay(vst(dds, blind = FALSE, nsub = 500))
vsd <- vst(dds, blind = FALSE, nsub = 500)
pcaData <-
  plotPCA(vsd,
          intgroup = "Group",
          returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```


## PCA plot for QC

PCA plot for the dataset that includes all libraries.

```{r pcaFull_comp, warnings=TRUE, message=FALSE}
rv <- rowVars(assay(vsd))
select <-
  order(rv, decreasing = TRUE)[seq_len(min(500, length(rv)))]
pca <- prcomp(t(assay(vsd)[select, ]))
percentVar <- pca$sdev ^ 2 / sum(pca$sdev ^ 2)
intgroup = "Group"
intgroup.df <- as.data.frame(colData(vsd)[, intgroup, drop = FALSE])
group <- if (length(intgroup) == 1) {
  factor(apply(intgroup.df, 1, paste, collapse = " : "))
}
d <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  intgroup.df,
  name = colnames(vsd)
)
```

plot PCA for components 1 and 2

```{r pcaFull_C1-C2, fig.cap="Figure 4: PCA plot for the first 2 principal components", fig.width=8, fig.height=5}
#nudge <- position_nudge(y = 0.5)
g <- ggplot(d, aes(PC1, PC2, color = Group)) +
  scale_shape_manual(values = 1:8) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  geom_point(size = 2, stroke = 2) +
  geom_text_repel(aes(label = name)) +
  xlab(paste("PC1", round(percentVar[1] * 100, 2), "% variance")) +
  ylab(paste("PC2", round(percentVar[2] * 100, 2), "% variance"))
g
#g
```

## Sample distance for QC


```{r distance, fig.cap="Figure 5: Euclidean distance between samples", fig.width=8, fig.height=5}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

```

## Set contrasts and find DE genes

```{r cotnrasts, warnings=TRUE, message=FALSE}
resultsNames(dds)
res.UndfvsDiff <- results(dds, contrast = c("Group", "Undf", "Diff"))
table(res.UndfvsDiff$padj < 0.05)
res.UndfvsDiff <- res.UndfvsDiff[order(res.UndfvsDiff$padj),]
res.UndfvsDiffdata <-
  merge(
    as.data.frame(res.UndfvsDiff),
    as.data.frame(counts(dds, normalized = TRUE)),
    by = "row.names",
    sort = FALSE
  )

```


## Creating gene lists

The gene lists have Ensembl gene-ID-version. We need `mirbase_id`,  `gene_biotype` and `external_gene_name` attached to make the results interpretable. 
So we will download metadata from ensembl using `biomaRt` package.

```{r annotations, warnings=TRUE, message=FALSE}
ensembl = useMart("ENSEMBL_MART_ENSEMBL")
ensembl = useDataset("mmusculus_gene_ensembl", mart = ensembl)
filterType <- "ensembl_gene_id_version"
filterValues <- rownames(cts)
attributeNames <- c('ensembl_gene_id_version',
                    'gene_biotype',
                    'mirbase_id',
                    'external_gene_name')
annot <- getBM(
  attributes = attributeNames,
  filters = filterType,
  values = filterValues,
  mart = ensembl
)
isDup <- duplicated(annot$ensembl_gene_id)
dup <- annot$ensembl_gene_id[isDup]
annot <- annot[!annot$ensembl_gene_id %in% dup, ] #this object will be saved and used later
```



## Write result tables

Here, we will attach the `mirbase_id`,  `gene_biotype` and `external_gene_name` downloaded in the previous section to the results table. We will also filter the the table to write:

1. DE list that have `padj` less than or equal to 0.05
2. DE list that have `padj` less than or equal to 0.05, and `log2FoldChange` greater than or equal to fold change of 1.5
3. DE list that have `padj` less than or equal to 0.05, and `log2FoldChange` less than or equal to fold change of -1.5
4. Full list of DE table without any filtering.

```{r output, warning=TRUE, message=FALSE}
names(res.UndfvsDiffdata)[1] <- "ensembl_gene_id_version"
res.UndfvsDiffdata <- merge(x = res.UndfvsDiffdata, y = annot, by = "ensembl_gene_id_version", all.x = TRUE)
res.UndfvsDiffdata <- res.UndfvsDiffdata[,c(1,(ncol(res.UndfvsDiffdata)-2):ncol(res.UndfvsDiffdata),2:(ncol(res.UndfvsDiffdata)-3))]
res.UndfvsDiffSig <- res.UndfvsDiffdata %>%
  filter(padj <= 0.05)
fc=1.5
log2fc=log(fc, base = 2)
neg.log2fc = log2fc * -1
res.UndfvsDiffSig.up <- res.UndfvsDiffdata %>%
  filter(padj <= 0.05 & log2FoldChange >= log2fc)
res.UndfvsDiffSig.dw <- res.UndfvsDiffdata %>%
  filter(padj <= 0.05 & log2FoldChange <= neg.log2fc)

write_delim(res.UndfvsDiffdata, file = "DESeq2_results_UndfvsDiff_full-table.tsv", delim = "\t")
write_delim(res.UndfvsDiffSig, file = "DESeq2_results_UndfvsDiff_adj.p-le-0.05.tsv", delim = "\t")
write_delim(res.UndfvsDiffSig.up, file = "DESeq2_results_UndfvsDiff_adj.p-le-0.05_fc-ge-1.5.tsv", delim = "\t")
write_delim(res.UndfvsDiffSig.dw, file = "DESeq2_results_UndfvsDiff_adj.p-le-0.05_fc-le-neg1.5.tsv", delim = "\t")
```

## Highly expressed genes for conditions


```{r expplot1}
exp <- res.UndfvsDiffdata[,c(1:4,11:ncol(res.UndfvsDiffdata))]
exp$gene <- paste0(exp$external_gene_name,"(",exp$gene_biotype,")")
renamed.exp <- exp[,c(13,2,4,5:12)]
renamed.exp.long <- melt(renamed.exp, id.vars=c('gene', 'gene_biotype', 'external_gene_name' ))
colnames(renamed.exp.long) <- c('gene', 'gene_biotype', 'external_gene_name', 'replicates', 'norm.expression')
renamed.exp.long$condition <- "NA"
renamed.exp.long$condition[which(str_detect(renamed.exp.long$replicates, "Dif_D6"))] <- "Diff"
renamed.exp.long$condition[which(str_detect(renamed.exp.long$replicates, "Undif_D2"))] <- "Undiff"
renamed.exp.long <- renamed.exp.long  %>% filter(norm.expression > 0) 
renamed.exp.long <- na.omit(renamed.exp.long)
# SOURCE: https://ourcodingclub.github.io/tutorials/dataviz-beautification/
theme_niwot <- function() {
  theme_bw() +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      axis.line.x = element_line(color = "black"),
      axis.line.y = element_line(color = "black"),
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.margin = unit(c(1, 1, 1, 1), units = , "cm"),
      plot.title = element_text(
        size = 18,
        vjust = 1,
        hjust = 0
      ),
      legend.text = element_text(size = 12),
      legend.title = element_blank(),
      legend.position = c(0.95, 0.15),
      legend.key = element_blank(),
      legend.background = element_rect(
        color = "black",
        fill = "transparent",
        size = 2,
        linetype = "blank"
      )
    )
}

```

## Normalized expression plots  {.tabset}

### Replicates

```{r vlnplot1, fig.cap="Figure 6A: Normalized expression of genes in undifferentiated and differentiated samples (replicate)", fig.width=8, fig.height=5}
A <- ggplot(data = renamed.exp.long,
       aes(x = replicates, y = norm.expression, fill = condition)) +
  geom_flat_violin(position = position_nudge(x = 0.2, y = 0),
                   alpha = 0.8,
                   trim = FALSE) +
  geom_point(
    aes(y = norm.expression, color = condition),
    position = position_jitter(width = 0.15),
    size = 1,
    alpha = 0.5
  ) +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               alpha = 0.8) + stat_summary(
                 fun = mean,
                 geom = "point",
                 shape = 23,
                 size = 2
               ) +
  labs(y = "\nNormalized Expression", x = NULL) +
  guides(fill = "none", color = "none") +
  scale_y_log10(label = comma)  +
  theme_niwot() + theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  ))
A
```

### Conditions
```{r vlnplot2, fig.cap="Figure 6B: Normalized expression of genes in undifferentiated and differentiated samples (condition)", fig.width=8, fig.height=5}
B <- ggplot(data = renamed.exp.long,
       aes(x = condition, y = norm.expression, fill = condition)) +
  geom_flat_violin(position = position_nudge(x = 0.2, y = 0),
                   alpha = 0.8,
                   trim = FALSE) +
  geom_point(
    aes(y = norm.expression, color = condition),
    position = position_jitter(width = 0.15),
    size = 1,
    alpha = 0.5
  ) +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               alpha = 0.8) + stat_summary(
                 fun = mean,
                 geom = "point",
                 shape = 23,
                 size = 2
               ) +
  labs(y = "\nNormalized Expression", x = NULL) +
  guides(fill = "none", color = "none") +
  scale_y_log10(label = comma)  +
  theme_niwot() + theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  ))
B
```

### Biotypes

```{r vlnplot3, fig.cap="Figure 6C: Normalized expression of genes in undifferentiated and differentiated samples (split based on gene biotype)", fig.width=8, fig.height=5}
C <- ggplot(data = renamed.exp.long,
       aes(x = gene_biotype, y = norm.expression, fill = condition)) +
  geom_flat_violin(position = position_nudge(x = 0.2, y = 0),
                   alpha = 0.8,
                   trim = FALSE) +
  geom_point(
    aes(y = norm.expression, color = condition),
    position = position_jitter(width = 0.15),
    size = 1,
    alpha = 0.5
  ) +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               alpha = 0.8) + stat_summary(
                 fun = mean,
                 geom = "point",
                 shape = 23,
                 size = 2
               ) +
  labs(y = "\nNormalized Expression", x = NULL) +
  guides(fill = "none", color = "none") +
  scale_y_log10(label = comma)  +
  theme_niwot() + theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  ))
C
```

## new section

## Volcano plots


```{r volcano2}
volcanoPlots2 <-
  function(res.se,
           string,
           first,
           second,
           color1,
           color2,
           color3,
           ChartTitle) {
    res.se <- res.se[order(res.se$padj), ]
    res.se <-
      rownames_to_column(as.data.frame(res.se[order(res.se$padj), ]))
    names(res.se)[1] <- "Gene"
    res.data <-
      merge(res.se,
            annot,
            by.x = "Gene",
            by.y = "ensembl_gene_id_version")
    res.data <- res.data %>% mutate_all(na_if, "")
    res.data <- res.data %>% mutate_all(na_if, " ")
    res.data <-
      res.data %>% mutate(gene_symbol = coalesce(external_gene_name, Gene))
    fc=1.5
log2fc=log(fc, base = 2)
neg.log2fc = log2fc * -1

    res.data$diffexpressed <- "other.genes"
    res.data$diffexpressed[res.data$log2FoldChange >= log2fc &
                             res.data$padj <= 0.05] <-
      paste("Higher expression in", first)
    res.data$diffexpressed[res.data$log2FoldChange <= neg.log2fc &
                             res.data$padj <= 0.05] <-
      paste("Higher expression in", second)
    res.data$delabel <- ""
    res.data$delabel[res.data$log2FoldChange >= log2fc
                     & res.data$padj <= 0.05
                     &
                       !is.na(res.data$padj)] <-
      res.data$gene_symbol[res.data$log2FoldChange >= log2fc
                           &
                             res.data$padj <= 0.05
                           &
                             !is.na(res.data$padj)]
    res.data$delabel[res.data$log2FoldChange <= neg.log2fc
                     & res.data$padj <= 0.05
                     &
                       !is.na(res.data$padj)] <-
      res.data$gene_symbol[res.data$log2FoldChange <= neg.log2fc
                           &
                             res.data$padj <= 0.05
                           &
                             !is.na(res.data$padj)]
    ggplot(res.data,
             aes(
               x = log2FoldChange,
               y = -log10(padj),
               col = diffexpressed,
               label = delabel
             )) +
      geom_point(alpha = 0.5) +
      xlim(-20, 20) +
      theme_classic() +
      scale_color_manual(name = "Expression", values = c(color1, color2, color3)) +
      # geom_text_repel(
      #   data = subset(res.data, padj <= 0.05),
      #   max.overlaps  = 15,
      #   show.legend = F,
      #   min.segment.length = Inf,
      #   seed = 42,
      #   box.padding = 0.5
      # ) +
      ggtitle(ChartTitle) +
      xlab(paste("log2 fold change")) +
      ylab("-log10 pvalue (adjusted)") +
      theme(legend.text.align = 0)
}
```
```{r volcano1, fig.cap="Figure 6: Volcano plot showing genes overexpressed in undifferentiated and differentiated states.", fig.width=8, fig.height=5}
g <- volcanoPlots2(
  res.UndfvsDiff,
  "UndfvsDiff",
  "Undf",
  "Diff",
  "tomato",
  "lightseagreen",
  "grey",
  ChartTitle = "Undifferentiated vs. Differentiated"
)
ggplotly(g)
#g
```
## Heatmap

Heatmap for the top 30 variable genes:

```{r heatmap, fig.cap="Figure 7: Heat map for top 30 variable small RNA genes", fig.width=8, fig.height=12}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 30)
mat  <- assay(vsd)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
mat2 <-    merge(mat,
          annot,
          by.x = 'row.names',
          by.y = "ensembl_gene_id_version")
mat2$gene <- paste0(mat2$external_gene_name," (",mat2$gene_biotype,")")
rownames(mat2) <- mat2[,'gene']
mat2 <- mat2[2:9]
heat_colors <- brewer.pal(9, "YlOrRd")
g <- pheatmap(
  mat2,
  color = heat_colors,
  main = "Top 30 variable small RNA genes",
  cluster_rows = T,
  cluster_cols  = T,
  show_rownames = T,
  border_color = NA,
  fontsize = 10,
  scale = "row",
  fontsize_row = 10
)
g
```
## MultiQC report:

MultiQC report is available at this [link](assets/multiqc_report.html){target="_blank"}


# Session Information

```{r sessioninfo}
sessionInfo()
```
